---
title: "Phenotypes associated with rare diseases"
author: "Jennifer HY Lin"
date: '2022-07-16'
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo = TRUE, message = FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(knitr)
```

### Initial data wrangling 
This data set was also derived and downloaded from Orphanet, but because it contained 37 columns with 112,243 rows originally, it took quite a long time to load on RStudio (or could be due to my laptop capacity...). It loaded relatively faster on Jupyter notebook from Anaconda, so I decided to clean it up first using Python there. Some columns were removed which reduced the number of columns from 37 to 13, while not changing any of the rows. The columns were also renamed to make it easier to read as shown below.

#### Source of dataset
* Orphadata: Free access data from Orphanet. 
* © INSERM 1999. Available on http://www.orphadata.org. Data version (XML data version).
* Dataset (.xml file) from http://www.orphadata.org/cgi-bin/epidemio.html
* Latest date of update for the dataset: 14/6/2022 (last accessed 24/7/2022)
* Creative Commons Attribution 4.0 International

```{r load file}
df <- read_csv("rare_disease_phenotypes.csv")
```

The abbreviation of HPO means human phenotype ontology.

```{r column names}
spec(df)
```

```{r full table}
df %>% View()
```
#### Exploratory data analysis
Since I wasn't intending for this project to grow into a TL;DR version (as most people would likely lose interests and willpower to read by then), I'd like to ask a question about the data set, in order to keep it at a reasonably short but informative length, which was what are the most common rare disorders along with their associated phenotypes?

So to answer it, let's observe the spread of the disorder groups and types by formulating a contingency table first.
```{r}
df_type <- df %>% 
  group_by(`Disorder group`,`Disorder type`) %>% 
  summarize(Number = n())
df_type
```
After a quick view on the column of "Disorder group", it mainly provided different disorder types a group label for each, which to a certain extent, was not necessary at this early stage of EDA. So I decided to remove this column for now from the contingency table, in order to focus solely on, "Disorder type" with number of counts (or times it appeared in the dataset).
```{r contingency table - disorder type & number}
df_type <- df %>% 
  group_by(`Disorder type`) %>% 
  summarize(Number = n())
df_type
```
Then to visualise this in a graphic way, a lollypop chart was built in a horizontal fashion, with different rare disorder types on the y-axis and number of each type on the x-axis.

```{r lollypop chart}
ggplot(data = df_type, aes(x = `Disorder type`, y = `Number`)) +
  geom_segment(aes(x = `Disorder type`, xend = `Disorder type`, y = 0, yend = `Number`), colour = "dark blue") +
  geom_point(colour = "dark green", size = 2, alpha = 0.6) +
  theme_light() +
  coord_flip() 
```
Two disorder types stood out the most, with Disease type appeared 57,920 times and Malformation syndrome at 37,634 times. To understand further what each of these two disorder types were, a direct reference was used (see note below) and according to the source of the dataset:

1.  The definition of "Disease" in the rare disease context is "a disorder with homogeneous therapeutic possibilities and an identified physiopathological mechanism...", one thing also worth noting was that this type did not include any developmental anomalies.

2.  For "Malformation syndrome", this was defined as, "A disorder resulting from a developmental anomaly involving more than one morphogenetic field. Malformative sequences and associations are included."

Reference: "Orphadata: Free access products description" - April 2020 <http://www.orphadata.org/cgi-bin/img/PDF/OrphadataFreeAccessProductsDescription.pdf> Version 2

To demonstrate this in a tabular form, with corresponding proportions of each disorder type in the dataset, the following codes were used:
```{r percentages of disorder type}
df1 <- df %>% 
  group_by(`Disorder type`) %>% 
  summarise(n = n()) %>% 
  mutate(prop = n/sum(n))
df1
```
The table was then rearranged with proportions in descending order (from highest to lowest). It also showed the top two were "Disease" (51.6%) and "Malformation syndrome" (33.5%).
```{r rearrage proportions}
df1 %>% arrange(desc(prop))
```
This was followed by checking out the distributions of HPO frequency to see which categories had the most and least number of counts.
```{r}
df_freq <- df %>% 
  count(`HPO frequency`) %>% 
  arrange(desc(n))
df_freq
```
Results for rare disorders with obligate or 100% frequency in patient's populations were then filtered, showing disorder type, HPO frequency and disorder name. Specifically, I wanted to find out the disorder names associated with the "Disease" disorder type with HPO frequency of "Obligate (100%)".
```{r }
df_freq_ob <- df %>% 
  filter(`Disorder type` == "Disease", `HPO frequency` == "Obligate (100%)") %>% 
  select(`Disorder type`, `HPO frequency`, `Disorder name`)
df_freq_ob
```
I'd then like to look into associated counts of appearance of each disorder name. When I cross-checked with the dataset table view, I've noted that the number of appearance of each disorder name is linked to the number of preferred HPO phenotype terms of each of these disorder types. 
```{r}
df2 <- df_freq_ob %>% count(`Disorder name`)
df2 %>% arrange(desc(n))
```
To show this, let's link preferred HPO terms to a disorder name such as this one, "Autosomal recessive complex spastic paraplegia due to Kennedy pathway dysfunction", which had the "Disease" disorder type with obligate or 100% HPO frequency.
```{r}
df %>% 
  filter(`Disorder type` == "Disease", `HPO frequency` == "Obligate (100%)", `Disorder name` == "Autosomal recessive complex spastic paraplegia due to Kennedy pathway dysfunction") %>% 
  select(`Disorder type`, `HPO frequency`, `Disorder name`, `Preferred HPO term`)
```
As shown in the dataframe above, under the column name, "Preferred HPO term", there were a total of ten different HPO phenotype terms associated with this particular rare disease with 100% HPO frequency within the patient population for this specific type of spastic paraplegia.

By using similar filtering method, we could quickly narrow down any particular rare disease of interest to find out specific phenotype or clinical features, along with associated HPO phenotype frequency, for further investigations.

For "Malformation syndrome" - let's look into associated phenotypes for it... *to be continued*