```{r}
library(tidyverse)
```

```{r}
df <- read_csv("Rare_diseases_drugs_FDA.csv")
```

```{r}
spec(df)
```

```{r}
df %>% View()
```

#Check the full view of the dataset

```{r}
select(df, `Generic Name`, `Trade Name`, `Date Designated`, `Orphan Designation`, `Approved Labeled Indication`, `Marketing Approval Date`, `Exclusivity Protected Indication * (Shown for approvals from Jan. 1, 2013, to the present)`, `Sponsor Country`)
```

```{r}
df %>% count(`Sponsor Country`)
```

```{r}
df1 = df %>% slice(-c(1058))
```

```{r}
df1 %>% count(`Sponsor Country`)
```

#Checked last row (index 1058) was only a caption for exclusivity period, so this was omitted from the selection #Count the number of sponsor countries involved again in rare disease drug developments (to remove last NA entry)

```{r}
ggplot(data = df1) +
  geom_bar(mapping = aes(x = `Sponsor Country`, fill = `Sponsor Country`)) +
  labs(title = "Countries involved in developing rare disease drugs", caption = "Dataset from US FDA: https://www.accessdata.fda.gov/scripts/opdlisting/oopd/index.cfm (last accessed: 1/7/2022)") +
  coord_flip()
```

#Plotting horizontal bar graph to show countries involved in rare disease drug developments (by counts)

```{r}
library(lubridate)
library(dplyr)
```

```{r}
df_uk <- df %>% 
  mutate(`Exclusivity End Date` = mdy(`Exclusivity End Date`), `Date Designated` = mdy(`Date Designated`), `Marketing Approval Date` = mdy(`Marketing Approval Date`)) %>%
  filter(year(`Exclusivity End Date`) > 2022 & `Sponsor Country` == "United Kingdom")
df_uk
```

#Filter rare disease drugs with Exclusivity End Date after 2022 & limited to UK as the sponsor country only

```{r}
install.packages("vistime")
```

```{r}
library(vistime)
```

#Use vistime to plot a timeline for rare disease drugs limited to the ones with exclusivity end date after 2022 from UK only

```{r}
gg_vistime(df_uk, col.event = "Trade Name", col.group = "Generic Name", col.start = "Date Designated", col.end = "Marketing Approval Date", title = "Timeline from designation to marketing approval")
```

```{r}
df_ir <- df %>%
  mutate(`Exclusivity End Date` = mdy(`Exclusivity End Date`), `Date Designated` = mdy(`Date Designated`), `Marketing Approval Date` = mdy(`Marketing Approval Date`)) %>%
  filter(year(`Exclusivity End Date`) > 2022, `Sponsor Country` == "Ireland")
df_ir
```

```{r}
gg_vistime(df_ir, col.event = "Trade Name", col.group = "Generic Name", col.start = "Date Designated", col.end = "Marketing Approval Date", title = "Timeline from designation to marketing approval")
```
#Plot timeline for rare disease drugs limited to the ones with exclusivity end date after 2022 from Ireland only
```{r}
df_us <- df %>%
  mutate(`Exclusivity End Date` = mdy(`Exclusivity End Date`), `Date Designated` = mdy(`Date Designated`), `Marketing Approval Date` = mdy(`Marketing Approval Date`)) %>%
  filter(year(`Exclusivity End Date`) == 2022, month(`Exclusivity End Date`) >= 1, month(`Exclusivity End Date`) < 7, `Sponsor Country` == "United States")
df_us
```
```{r}
gg_vistime(df_us, col.event = "Trade Name", col.group = "Generic Name", col.start = "Date Designated", col.end = "Marketing Approval Date", title = "Timeline from designation to marketing approval")
```
#Limiting rare disease drugs from US with exclusivity end date from January to June 2022 only (shown in a timeline)